The Evolution of Phenotypically Invariant Genetic Networks

### 2017-08-29 kalman_walk_paper.tex
F2 hybrid vigor calc
hybrid vigor
edited system description, qgen, examples, speciation and more.

\newcommand{\calA}{\mathcal{A}}
\newcommand{\calH}{\mathcal{H}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

@@

% Other title ideas:
%
% More than one way to grow a cat: regulatory network drift and speciation
% 
% Rapid speciation despite conservation of phenotype
%
% More than one way to grow a cat: regulatory network redundancy and speciation
% 
% Neutral network drift leads to incompatibilities on the same time scale as genetic drift
% 
% More than one way to grow a cat: can network drift lead to speciation?
% 
% How fast does network drift create incompatibilities?
% 
% Beyond the snowball: a quantitative model of incompatibility accumulation
% 
% Systems drift and speciation: more than one way to grow a cat
% 
% An explicit model of neutral regulatory network evolution, with applications
% to the rate of accumulation of hybrid incompatibility
% 
% Evolutionary conservation of phenotype does not entail conservation of the underlying
% molecular mechanism leading to rapid speciation
%
% The evolution of phenotype-invariant gene networks rapidly leads to hybrid incompatibiliy
%
% Evolutionary network rewiring can rapidly lead to hybrid incompatibility despite
% phenotypic conservation
%
% Hybrid incompatibilities can evolve rapidly due to developmental systems drift
%
% Evolutionary systems theory and speciation

@@

\section*{Gene Networks as Linear Dynamical Systems}

\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
Rather than keep making comments like "(or some other system)",
we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}

%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. 
%   Thus we say that some function $\phi(t)$ is a phenotype where, 
%  \begin{align}
%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
%  \end{align}
%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity. 
%
%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
%
%  We describe the impulse response as, 
%  \begin{align}
%    h(t) = C e^{A t} B
%  \end{align}
%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
%
%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
%  \begin{align}
%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
%  \end{align}
%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
%}
\js{Will rewrite this section again.} Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. 
We define an organism's ``phenotype'' to be the temporal dynamics of molecular concentrations directly relevant to survival, 
and denote by $\phi_k(t)$ the concentration of the $k^\text{th}$ element of the phenotype at time $t$, given an input; for instance $\phi_k(t)$ could describe the concentrations of a particular enzyme.
These dynamics are the result of the interconnections of 
a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) 
and an environmental input vector $u(t)$, and the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}. 
Such a system $\mathcal{S}$, is composed of two equations,
  \begin{align}
    \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
  \end{align}
As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the system's \emph{external dynamics}.
The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the system's current state (the kryptotype), the system architecture $(A,B,C)$ -- such as the wiring of a gene regulatory network, and the environmental input $u(t)$ to the system. The architecture of $A$ determines exactly how the internal molecular concentrations regulate one another's expression as $A_{ij}$ is the magnitude at which transcription factor $\kappa_j(t)$ regulates transcription factor $\kappa_i(t)$. If $A_{ij} > 0$, we say that $\kappa_j$ upregulate $\kappa_i$, and if $A_{ij} < 0$, we say that $\kappa_j$ down-regulates $\kappa_i$. Furthermore, we interpret the $i^\text{th}$ row of $A$ to be the \emph{cis}-regulatory element (or promoter) for gene $i$.  
The architecture of $B$ determines how the system processes the environmental intput, and the structure of $C$ converts the internal dynamics to external dynamics. That is if $\mathcal{S}$ is a metabolic system, $C_{ij}$ is the amount at which the $j$th input metabolite $u_j$ affects the production of the $i$th enzyme concentration.

\plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}
Thus the phenotype is simply a convolution of the system organization and the environment, 
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
where we refer to $h(t) := Ce^{A t}B$ as the system's \emph{impulse response}. 
%We will now lay out the model in more general terms.
%Suppose that the \emph{internal state} of the system
%is parameterized by the concentrations of a collection of $n$ molecular species,
%$S_1, \ldots, S_n$,
%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
%There are also $m$ ``input'' species, whose concentrations are determined
%exogenously to the system,
%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
%and $\ell$ ``output'' species, whose concentrations are denoted
%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
%The output is merely a linear function of the internal state:
%\begin{align*}
%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
%\end{align*}
%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
%as it is ``hidden'' from direct selection.
%The rate at which the $i^\text{th}$ species is produced
%is a weighted sum of the concentrations of the other species
%as well as the input:
%\begin{align*}
%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
%\end{align*}
%In matrix notation, this is written more concisely as
%\begin{align} \label{eqn:lti_system}
%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
%    \phi(t) &= C \kappa(t) .
%\end{align}

  \section*{Gene Network Drift}

@@

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal system (a gene network) 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes 
-- has an identical impulse response and therefore an identical phenotype given an identical input $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
\plr{what are we calling these -- not "biological system" -- that's too broad}\js{how about ``molecular systems''?}
Two molecular systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

@@

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore, if a system is minimal, the set of all phenotypically identical systems $\mathcal{S}(V)$ can be parameterized as,
    \begin{align}
      \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  \end{align}
  where $V$ is any invertible matrix and it's elements are free parameters. 
More generally,
we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
  \calA_n(A_0) 
      &= \left\{
          A : C e^{At} B = C e^{A_0 t} B \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
          A : C A^k B = C A_0^k B \; \text{for}\; 1 \le k \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

\plr{should say more about this decomposition here.}
Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$. This set can be completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined. Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.

@@

%\plr{This is not part of the definition of $\mathcal{I}$:}
%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
%%\plr{don't need these to be big equations.}
%%\begin{align*}
%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
%%  \mathcal{I} &= 0
%%\end{align*}
%
%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
%
%  \begin{figure}[H]
%    \centering
%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
%  \end{figure}
%  }
%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
%%       DMI examples...
%%       \begin{figure}[H]
%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
%%       \end{figure}
%%     \end{example}
%
%%     \begin{example}[Not all Networks can Host Incompatibilities]
%%       convex sets cant have DMIs
%
%%       \begin{align*}
%%         h(t) = 2 e^{- \theta t}
%%       \end{align*}
%%       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
%%     \end{example}
%
%
%\plri{Missing: a statement about when there are neutral directions, and how many.}
%
%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}

>>>>

\newcommand{\calA}{\mathcal{S}}
\newcommand{\calH}{\mathcal{H}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

@@

% Other title ideas:
%
% More than one way to grow a cat: speciation despite conservation of phenotype
%
% More than one way to grow a cat: regulatory network drift and speciation
%
% Rapid speciation despite conservation of phenotype
%
% More than one way to grow a cat: regulatory network redundancy and speciation
%
% Neutral network drift leads to incompatibilities on the same time scale as genetic drift
%
% More than one way to grow a cat: can network drift lead to speciation?
%
% How fast does network drift create incompatibilities?
%
% Beyond the snowball: a quantitative model of incompatibility accumulation
%
% Systems drift and speciation: more than one way to grow a cat
%
% An explicit model of neutral regulatory network evolution, with applications
% to the rate of accumulation of hybrid incompatibility
%
% Evolutionary conservation of phenotype does not entail conservation of the underlying
% molecular mechanism leading to rapid speciation
%
% The evolution of phenotype-invariant gene networks rapidly leads to hybrid incompatibiliy
%
% Evolutionary network rewiring can rapidly lead to hybrid incompatibility despite
% phenotypic conservation
%
% Hybrid incompatibilities can evolve rapidly due to developmental systems drift
%
% Evolutionary systems theory and speciation

@@

\section*{Gene Networks as Linear Dynamical Systems}
%
\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
Rather than keep making comments like "(or some other system)",
we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}
%
%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. 
%   Thus we say that some function $\phi(t)$ is a phenotype where, 
%  \begin{align}
%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
%  \end{align}
%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity. 
%
%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
%
%  We describe the impulse response as, 
%  \begin{align}
%    h(t) = C e^{A t} B
%  \end{align}
%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
%
%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
%  \begin{align}
%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
%  \end{align}
%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
%}
Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.

Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regualtory network), given by the triple $(A,B,C)$,
and by its environment $u(t)$.

Such a biological system $\mathcal{S}$ is given by,
  \begin{align}
    \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
  \end{align}

Generally $A$ is any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
%
Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
%The $i^\text{th}$ row of $A$ contains the regulatory logic pertaining to gene $i$.
%
As such, $A_{ij}$ is the magnitude at which transcription factor $\kappa_j(t)$ regulates transcription factor $\kappa_i(t)$, and if $A_{ij} > 0$, we say that $\kappa_j$ upregulates $\kappa_i$. If $A_{ij} < 0$, we say that $\kappa_j$ down-regulates $\kappa_i$.

The form of $B$ determines precisely how the environment influences the organism -- that is $B$ filters and translates the input to the system. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.
\emph{E.g.} for a metabolic system, $C_{ij}$ is the amount the $j$th metabolite $u_j$ affects the production of the $i$th enzyme.
%

  %where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, 
  %and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%
Furthermore, whereas the phenotype is a subset of molecules \emph{visible} to selection, the \emph{kryptotype} includes the molecular dynamics \emph{hidden} from selection. That is $\kappa(t)$ is a vector of the system's molecular concentrations at time $t$, directly, indrectly, and irrelevant to survival.
%
%The structure of $B$ and $C$ determines how systems process their environment and convert krypto- to phenotypes, respectively. 
%
\plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}

Finally, we can write the phenotype as a convolution of the system organization and the environment,
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
where we refer to $h(t) := Ce^{A t}B$ as the system's \emph{impulse response}. 
%We will now lay out the model in more general terms.
%Suppose that the \emph{internal state} of the system
%is parameterized by the concentrations of a collection of $n$ molecular species,
%$S_1, \ldots, S_n$,
%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
%There are also $m$ ``input'' species, whose concentrations are determined
%exogenously to the system,
%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
%and $\ell$ ``output'' species, whose concentrations are denoted
%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
%The output is merely a linear function of the internal state:
%\begin{align*}
%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
%\end{align*}
%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
%as it is ``hidden'' from direct selection.
%The rate at which the $i^\text{th}$ species is produced
%is a weighted sum of the concentrations of the other species
%as well as the input:
%\begin{align*}
%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
%\end{align*}
%In matrix notation, this is written more concisely as
%\begin{align} \label{eqn:lti_system}
%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
%    \phi(t) &= C \kappa(t) .
%\end{align}
 \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}
    Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}. Here we consider a simplified model of oscillating gene transcription.

    Suppose gene-2 up-regulates the transcription of gene-1 and that gene-1 down-regulates gene-2 with equal magnitudes, whose concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$. Furthermore, suppose that only the dynamics of gene-1 are consequential to the cell cycle (perhaps the amount of gene-1 activates another downstream gene network). Lastly suppose that the production of both genes is stimulated by an impulse of a molecule present immediately after division.

    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
    \end{align*}
    where $\dot{\kappa}$ denotes the time derivative. The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time. If we record the regulatory coefficients in the matrix
    \begin{align*}
      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
    \end{align*}
    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
      \begin{align*}
      \dot{\kappa} &= A \kappa(t) + B u(t) .
      \end{align*}
      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
      \begin{align*}
      \phi(t) &= C \kappa (t)
      \end{align*}
      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. 
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}

  \section*{Phenotype-equivalent gene networks}

@@

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal system (a gene network) 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes 
-- has an identical impulse response and therefore an identical phenotype given an identical input $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
\plr{what are we calling these -- not "biological system" -- that's too broad}\js{I see your point, but I don't want to bog down the paper by constantly qualifying statements -- it seemsnatural to refer to biological systems, since these are, afterall dynamical systems.}
Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

@@

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore, if a system is minimal, the set of all phenotypically identical systems $\mathcal{S}(V)$ can be parameterized as,
    \begin{align} \label{eq:min_param}
      \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  \end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by \ref{eq:min_param}, but they are ecanpuslated by \ref{eqn:equivalence}, below. 
More generally,
we denote by $\calA_n(\mathcal{S}_0)$ the set of all $n$-dimensional systems equivalent to $\mathcal{S}_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\mathcal{S}_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

Two systems, even if in different dimensions, can have identical phenotypes if they are in $\calA_n$. Further, there is no unique triple $(A,B,C)$ for the mapping $u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is \emph{all} gene regulatory networks can drift/rewire. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}.

@@

%\plr{This is not part of the definition of $\mathcal{I}$:}
%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
%%\plr{don't need these to be big equations.}
%%\begin{align*}
%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
%%  \mathcal{I} &= 0
%%\end{align*}
%
%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
%
%  \begin{figure}[H]
%    \centering
%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
%  \end{figure}
%  }
%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
%%       DMI examples...
%%       \begin{figure}[H]
%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
%%       \end{figure}
%%     \end{example}
%
%%     \begin{example}[Not all Networks can Host Incompatibilities]
%%       convex sets cant have DMIs
%
%%       \begin{align*}
%%         h(t) = 2 e^{- \theta t}
%%       \end{align*}
%%       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
%%     \end{example}
%
%
%\plri{Missing: a statement about when there are neutral directions, and how many.}
%
%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
%
%
 \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}

%\plr{this is same as above, so reference instead of writing out again?}
%    Let
%    \begin{align*}
%      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
%    \end{align*}
%    with $V(\tau)$ preserving both $B$ and $C$, then
    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\end{tikzpicture}
    \caption{The set of all phenotype-equivalent cell cycle control networks, $A(\tau)$.
      $A(\tau) = \left\{ \calA_2 : B = B_0,\ C = C_0\right\}$.}
    \end{figure}
 % \end{example}

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
    \caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom). Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
  \end{figure}
\end{example}

\paragraph{Sexual reproduction and speciation}
A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks. 
Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
